---
format: gfm
engine: knitr
execute:
  daemon: false
---

<!-- --8<-- [start:software] -->

- Unix terminal (e.g., `bash`)
- [Python environment with taxpasta](/#install)
- [R](https://rstudio-education.github.io/hopr/starting.html#how-to-download-and-install-r)
- Package dependencies

    ```R
    install.packages(c("readr", "dplyr"), dependencies = TRUE)
    ```

<!-- --8<-- [end:software] -->

<!-- --8<-- [start:raw-motus] -->

We can try loading a mOTUs profile into R using the common table reading function `read_tsv()` from the `readr` package with default arguments.

```{r}
requireNamespace("readr")
profile_motus <- readr::read_tsv("2612_pe-ERR5766176-db_mOTU.out")
```

You can see we immediately hit an error, as there is a 'comment' line at the top of the mOTUs profile with information on how the profile was generated.

While such a comment is very useful for reproducibility, to load this we have to instead add extra options to the function, which makes loading the table less than smooth for downstream analyses.

```{r}
profile_motus <- readr::read_tsv("2612_pe-ERR5766176-db_mOTU.out", comment = "#")
```

However, once again we hit another error: the column headers are _also_ specified as a comment line...  Instead we can try to skip the first two lines entirely.

```{r}
profile_motus <- readr::read_tsv("2612_pe-ERR5766176-db_mOTU.out", skip = 2)
profile_motus
```

<!-- --8<-- [end:raw-motus] -->

<!-- --8<-- [start:raw-kraken2] -->

```{r}
profile_kraken2 <- readr::read_tsv("2612_pe-ERR5766176-db1.kraken2.report.txt")
profile_kraken2
```

This doesn't fail to load but unfortunately the column headers look a bit weird. It seems the Kraken2 file does not include a column header! In this case we have to specify these ourselves.

```{r}
profile_kraken2 <- readr::read_tsv(
    "2612_pe-ERR5766176-db1.kraken2.report.txt",
    col_names = c(
        "percent",
        "clade_assigned_reads",
        "direct_assigned_reads",
        "taxonomy_lvl",
        "taxonomy_id",
        "name"
    )
)
profile_kraken2
```

<!-- --8<-- [end:raw-kraken2] -->

<!-- --8<-- [start:outer-join] -->

In the tidyverse flavour of R, we can do this with the `full_join` function of
the `dplyr` package. This form of joining tables includes all rows both from the
left and right table in the resulting table.

```{r}
#| error: true
requireNamespace("dplyr")
dplyr::full_join(profile_motus, profile_kraken2)
```

The error `by must be supplied when x and y have no common variables` occurs
because the column names are not the same between the two tables for the
different profilers' outputs. We need to specify which column of the left table
should be joined with what column of the right table.

```{r}
raw_merged_table <- dplyr::full_join(profile_motus, profile_kraken2, by = c("NCBI_tax_id" = "taxonomy_id"))
raw_merged_table
```

<!-- --8<-- [end:outer-join] -->

<!-- --8<-- [start:std-kraken2] -->

Now let’s try to load the taxpasta standardised Kraken2 result into
R again.

```{r}
profile_kraken2_std <- readr::read_tsv("2612_pe-ERR5766176-db1_kraken2.tsv")
profile_kraken2_std
```

<!-- --8<-- [end:std-kraken2] -->

<!-- --8<-- [start:merge-motus] -->

Once again, let's try loading the standardised and merged mOTUs result into R.

```{r}
profile_motus_merged <- readr::read_tsv("dbMOTUs_motus.tsv")
profile_motus_merged
```

<!-- --8<-- [end:merge-motus] -->
